---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library("shapley")
```
# shapley

[![CRAN Version](https://www.r-pkg.org/badges/version/shapley)](https://CRAN.R-project.org/package=shapley)
[![Build Status](https://travis-ci.org/elbersb/shapley.svg?branch=master)](https://travis-ci.org/elbersb/shapley)
[![Coverage status](https://codecov.io/gh/elbersb/shapley/branch/master/graph/badge.svg)](https://codecov.io/github/elbersb/shapley?branch=master)

The Shapley value is a concept from game theory that quantifies how much each player contributes
to the game outcome (Shapley 1953). The concept, however, has many more use cases: it provides
a method to quantify the importance of predictors in regression analysis or machine learning models,
and can be used in a wide variety of decomposition problems (Shorrocks 2013).
Most implementations focus on one narrow use case, although the algorithm for the Shapley
value decomposition is always the same -- it is just the concrete value function that varies.
This package provides a simple algorithm for the Shapley value decomposition.

## Installation

``` r
devtools::install_github("elbersb/shapley")
```

## Usage

The package provides a `shapley` function that takes two arguments: the value function and a
vector of factor names. The value function needs to be an R function that takes one argument.
The `shapley` function will call the value function repeatedly, each time with a different
set of factors.

For a very simple example, consider that an outcome is determined by two factors "A" and "B", which
contribute 1 and 2, respectively. The factors are linearly additive, which makes the use of Shapley
value decomposition unnecessary, but this allows us to validate the results.
The value function is thus defined as:

``` {r}
simple <- function(factors = c()) {
    value <- 0
    if ("A" %in% factors) value <- value + 1
    if ("B" %in% factors) value <- value + 2
    return(value)
}
```

We now supply the value function to `shapley`, along with the factor names:

``` {r}
shapley(simple, c("A", "B"), silent = TRUE)
```

As expected, the marginal contributions of the two factors are 1 and 2, respectively.
For factor "A", the underlying computation is:

``` {r}
1/2 * (simple("A") - simple()) + 1/2 * (simple(c("A", "B")) - simple("B"))
```

### Example 1: Game theory

For this example (taken from [Wikipedia](https://en.wikipedia.org/wiki/Shapley_value#Glove_game)),
consider three players. Players 1 and 2 supply right-hand gloves,
while Player 3 supplies a left-hand glove. The game is only successful if players with both types
of gloves enter into a coalition.
We thus define the value function as 1 if pairs `{1,3}`, `{2,3}` or `{1,2,3}`
are formed, and 0 otherwise.

In R, we can define the value function as follows:

``` {r}
glove <- function(factors) {
    if (length(factors) > 1 & 3 %in% factors) return(1)
    return(0)
}
```

To compute the marginal contributions of each player, use:

``` {r}
shapley(glove, c(1, 2, 3), silent = TRUE)
```

### Example 2: Relative importance of predictors

Consider this simple regression model and its R<sup>2</sup>:

```{r}
model <- lm(mpg ~ wt + qsec + am, data = mtcars)
summary(model)$r.squared
```

The Shapley value decomposition allows us to determine how much each predictor
contributes to the R<sup>2</sup>. To do this, we need to define the value function in a way
that it runs the regression with the appropriate subset of predictors. It should return 0 when
there are no predictors:

``` {r}
reg <- function(factors) {
    if (length(factors) == 0) return(0)
    formula <- paste0("mpg ~ ", paste(factors, collapse = "+"))
    m <- lm(formula, data = mtcars)
    summary(m)$r.squared
}

# test - should be the same as above:
reg(c("wt", "qsec", "am"))
```

```{r}
shapley(reg, c("wt", "qsec", "am"), silent = TRUE)
```

Note that there are many packages (e.g., [relaimpo](https://cran.r-project.org/package=relaimpo))
that provide this functionality specifically for regression analysis.

## References

Shapley, L. S. (1953). A value for n-person games.
    Contributions to the Theory of Games, 2(28), 307-317.

Shorrocks, A. F. (2013).
    Decomposition procedures for distributional analysis:
    a unified framework based on the Shapley value. Journal of Economic Inequality, 1-28.
